RL-Musician: A Tool for Music Composition with Reinforcement Learning

### 2020-01-26 docs/paper/paper.tex
docs: Write conclusion

\date{January 11, 2020}

@@

\begin{table}[h!]
	\caption{Scores of sample pieces}
	\label{table:scores}
	\begin{center}
		\begin{tabular}{|l|c|c|c|c|}
			\hline
			& Baseline & Agent & Human & Maximum\\
			\hline
			Absence of looped pitches & & & 0 & 0\\
			\hline
			Absence of pitch or pitch class clashes & & & -0.643 & 0\\
			\hline
			Motions & & & 0.1 & 1\\
			\hline
			Correlation between lines & & & 0.278 & 1\\
			\hline
			Explicity of climax & & & 3 & 3\\
			\hline
			Number of skips & & & 1 & 1\\
			\hline
			Total reward & & & 3.735 & 6\\
			\hline
		\end{tabular}
	\end{center}
\end{table}

It can be seen that .
%different runs are variations of the same piece.
%pitch class clashes occur.

As for subjective evaluation of music quality, it is not correct to compare the selected piece created by an agent with the piece created by the author. The problem is that reward encourages to follow counterpointal rules but, from a modern listener point of view, breaking these rules does not necessarily sound unaesthetically. As a result, subjective perception of a resulting piece is not aligned with objective function used during optimization.

Table \ref{table:piece} contains the created by an agent piece for which above scores are granted.

\begin{table}[h!]
	\caption{A piece composed by an agent}
	\label{table:piece}
	\begin{center}
		\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

@@

>>>>

\date{January 26, 2020}

@@

\begin{table}[h!]
	\caption{Scores of sample pieces}
	\label{table:scores}
	\begin{center}
		\begin{tabular}{|l|c|c|c|c|}
			\hline
			& Baseline & Agent & Human & Maximum\\
			\hline
			Absence of looped pitches & -1 & 0 & 0 & 0\\
			\hline
			Absence of pitch or pitch class clashes & -0.429 & -0.429 & -0.643 & 0\\
			\hline
			Motion & -0.1 & 0.367 & 0.1 & 1\\
			\hline
			Correlation between lines & 0.552 & 0.599 & 0.278 & 1\\
			\hline
			Explicity of climax & 0.9 & 1.65 & 3 & 3\\
			\hline
			Number of skips & 0 & 1 & 1 & 1\\
			\hline
			Total reward & -0.076 & 3.187 & 3.735 & 6\\
			\hline
		\end{tabular}
	\end{center}
\end{table}

It can be seen that the agent's piece is better than the manually created piece by all metrics except explicity of climax. Since climax is higher than start note, both lines are assumed to have upward motion but the agent wrongly trades off climax score for contrary motion and lower lines correlation.

Table \ref{table:piece} contains the created by the agent piece for which above scores are granted.

\begin{table}[h!]
	\caption{A piece composed by an agent}
	\label{table:piece}
	\begin{center}
		\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
			\hline
			& & & & & & & C6 & & & & & & & & \\
			\hline
			& & & & & & B5 & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & A5 & & & & A5 & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & G5 & & & & G5 & & G5 & G5 & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & F5 & & & & & & & & & & & & \\
			\hline
			& & E5 & & & & & & & & & & E5 & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& D5 & & & & & & & & & & & & & D5 & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			C5 & & & & & & & & & & & & & C5 & & C5 \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			G4 & & & & & & & & & & & & & & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& F4 & & & & & & & & & & & & & & \\
			\hline
			& & E4 & & & E4 & & E4 & & & E4 & & & E4 & & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & D4 & & & D4 & & D4 & & & D4 & & & D4 & \\
			\hline
			& & & & & & & & & & & & & & & \\
			\hline
			& & & & C4 & & & & & C4 & & & C4 & & & C4 \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

As for subjective evaluation of music quality, it is not correct to compare the selected piece created by the agent with the piece created by the author. The problem is that reward encourages to follow counterpointal rules but, from a modern listener point of view, breaking these rules does not necessarily sound unaesthetically. That being said, subjective perception of a resulting piece is not aligned with fitness function used during optimization. However, it must be noted that the piece shown above looks like overfitted to the given fitness function. It is not a problem associated with training, it is a flaw in problem setup.

Numerous experiments revealed another important issue. Actually, different runs of a trained agent produce variations of the same piece. Although there might be significant differences in scores (in particular, because some scoring functions are discrete), there are no significant differences from music point of view.

@@

The list of suggested further improvements is as follows:
\begin{itemize}
	\item \textbf{Reflect more knowledge from music theory in filtering rules and evaluational rules.} For example, overlapping motion should be discouraged. Also downward skips can be penalized in order to have upward skips before climax and downward stepwise motion after it.
	\item \textbf{Promote variety in fitness function.} Now, fitness function just averages rewards for several episodes. However, a new term can be added to this average. This term must reflect how different created pieces are from each other.
	\item \textbf{Revise feature representation for actor model.} Probably, procedure of next measure generation should be changed. Currently, it is assumed that shallow neural network is able to act based on relatively raw feature representation. To make it less raw, some features inspired by music theory may be engineered.
	\item \textbf{Abstract away all rules and focus only on essentials of music.} Tonal music introduces attraction between pitches. Depending on this attraction and degree of consonance/dissonance, tension increases or dissipates. Moments of time when level of tension changes may be rhythmically accented or unaccented. Close but not exact alignment between rhythmical accent and resolution results in aesthetic and interesting music. Taking into account voice leading constraints, optimization problem can be set up based on above considerations.
\end{itemize}
